Reproducibility of SNV-calling in multiple sequencing runs from single tumors

### 2015-09-30 PAPER/paper.tex
1-2/13 revision

We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol. For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192. Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the number of SNVs found in a tumor) between samples. This result may be in part due to known mutational hotspots in some tumors \citep{Karen}. Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation (a site with more than one alternate allele, in addition to the reference) may appear fixed or completely absent. In general, we would like to know how often these errors occur. One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation. However, this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates. Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.

>>>>

We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol. For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192. Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor. This result may be in part due to known mutational hotspots in some tumors \citep{Karen}. Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation, defined as a mutation that is not fixed and therefor is not present in all tumor cells, may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing. Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen. In general, we would like to know how often these errors occur. One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation. However, this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates. Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
